A Survey of Scientific Workflow Management Systems

### 2014-04-07 paper.tex
new bib entries citing recent publications

%Workflows considered:
% MOTEUR2
% Taverna
% Swift
% Pegasus
% Triana
% Galaxy
% Makeflow
% Kepler/Ptolemy
% Dryad
% Science pipes
% Spark
% Linda
% DaGue

@@

\vspace*{0.5cm}\hspace{1cm} Scientific Workflow Environments (SWE from here on)
are, in principle similar to the traditional programming environments
(programming language, interpreter or compiler, and runtime) albeit with a
larger scale of data and a higher abstraction of computation. The similarity is
such that scripting languages are sometimes advocated for programming at this
level. However, traditional scripts do not deal with parallelism and dataflow
aspects of workflow paradigm that makes them a limited contenders. 
%
%Add more content here to beef up the context.
%
The goals of a workflow management system for scientific computing are as
follows:

\begin{enumerate}
\item \textbf{User Interaction.} Adaptation to new methods of computing is not
easy for domain science users. Such users often need special assistance in
order to translate their computation requirements into concrete computing
entities.  SWEs are an ideal platform enabling users to perform many tasks
conveniently.  These tasks include composing and marshalling of execution,
browsing and selecting data, visualizing results. Furthermore, dissemination of
methods and results to communities and collaborators are some of the activities
that SWEs make convenient for end-users.

@@

@@

\item \textbf{Interface to Distributed Computing Infrastructures (DCIs).}
Building SWE's interfaces to distributed systems is a complex task. The
distributed nature of resource utilization and asynchronous execution of
workflow requires a set of interdependent parameters to be configured
correctly. Further complexity is added because of complex dataflow
requirements, interactions with diverse tools and linking with legacy
application codes.

%
\item \textbf{Expressiveness.} Programming languages are inherently control
flow oriented where data flow is a secondary aspect. In fact, the
programmer must explicitly make sure that correct the inputs are available
to instructions and are in the right format. A program proceeds as the
execution of its statements from beginning to end and completes its
execution in a predefined sequence. However, in SWEs, the flow is often
determined by the data availability for execution. This model of
computation is known as dataflow computing. A workflow thus does not follow
a fixed order of execution and different execution sequences are possible
which are dominated by dataflow. The unit of work in SWEs is at a higher
level of granularity. A unit of execution is coarse-grained, and often
termed as ``processor'', ``task'', ``service'' or ``actor'' in the SWE
community. This notion of high-level granularity drives the depth to which
a workflow can be expressed in a given language form. In the rest of this
paper, this unit of work will be referred to as \textit{processor}. On the
other hand, a unit of execution in most programming languages is a much
simpler and finer executable statement that accomplishes a small
computational task. This presents a whole new set of requirements for
expressing the processors and how they should interact and operate upon
data.

@@

%
\begin{figure}[htb]
\begin{center}
\includegraphics[width=\linewidth]{figures/workflow_interfaces_and_language}
\caption{A generic SWE scheme showing the relationship and interaction among important components}
\label{fig:wfms}
\end{center}
\end{figure}
%
SWEs are used as tools to benefit from the underlying large-scale computing
infrastructure such as clusters, supercomputers, clouds, grids. Such
environments are in use today for production level work in different scientific
domains worldwide. They do the basic task of linking inter-related processors
of a multi-staged scientific computation. However, they differ in the features
they provide based upon the scientific domain they offer service for, and the
computational model they follow. Often, one of the main reasons for such
difference is the specialized components and requirements for a particular
domain. Owing to this fact, a given workflow system dominates a particular area
of science (eg. astronomy, biology, etc.). Nonetheless, there exist a plethora
of SWEs suitable for general purpose workflow style computations.
% Ketanmark

@@

\section{Objectives and Organization}
The present work analyzes contemporary SWEs in order to find the capabilities
and key features in terms of the characteristics described in the introduction
part of this paper. The rest of this paper is organized as follows: Section
\ref{sec:overview} presents an overview of the key SWEs studied in this work.
Section \ref{sec:scal} discusses the optimization features and resulting
scalable performance as observed in the SWEs studied. Section \ref{sec:data}
studies the data management and description aspects of the SWEs. Section
\ref{sec:inter} presents a study of SWEs approaches to interface the
large-scale computing infrastructures. Section \ref{sec:lang} details the
language features and expressiveness of the SWEs studied. Section
\ref{sec:usable} discusses the usability aspects of the SWEs.  Section
\ref{sec:sum} presents a summary and a classification of all the aspects of
SWEs studied in this work. Throughout the paper, we also present tables
summarizing the aspects of SWEs considered in respective sections. It was not
possible to obtain an exhaustive information about all the SWEs considered in
this work since we rely on the peer reviewed publications for information on a
given SWE. Some features, such as extendability of SWE strongly depend on the
extension points designed in the code; consequently, we consider it present in
only those SWEs that have a documented description or demonstration of the
given feature.

@@

@@

%

\paragraph{Swift/Swift Script} Swift \cite{swift} is a script-based scientific
workflow environment developed at the University of Chicago. A Swift workflow
specified in the SwiftScript scripting language is translated to an
intermediate XML-based representation and enacted by the Karajan workflow
enactment engine. The SwiftScript language is proposed as a workflow language
founded on a scripting approach programmers are familiar with. To make the
language adapted to grids, the inventors of SwiftScript generalized the use of
futures \cite{Halstead85multilisp:a} for every variables defined in the script.
Futures are non-blocking assignment variables, using a proxy to ensure
immediate execution of assignment and performing lazy blocking on variable
value accesses only. The systematic use of futures for every variables in the
language makes it completely data-driven: the execution progresses
asynchronously as long as a data access is not blocking. The availability of
data enables blocked thread to restart execution. The beauty of futures is to
make this process completely hidden from the programmer who can use a scripting
language he is familiar with without ever manipulating explicit parallel
programming constructs. Swift facilitates a highly asynchronous workflow
execution. Swift supports advanced workflow constructs such as parallel loops
and conditionals. Furthermore Swift supports processing of explicitly defined
arrays of data. Swift workflows take advantage of POSIX-style shared file
system to reuse the data produced by other processors.

>>>>

%Workflows considered:
% MOTEUR2
% Taverna
% Swift
% Pegasus
% Triana
% Galaxy
% Makeflow
% Kepler/Ptolemy
% Dryad
% Science pipes
% Spark
% Linda
% DaGue
% Wings

@@

\vspace*{0.5cm}\hspace{1cm} Scientific Workflow Environments (SWE from here on)
are, in principle similar to the traditional programming environments
(programming language, interpreter or compiler, and runtime) albeit with a
larger scale of data and a higher abstraction of computation. The similarity is
such that scripting languages are sometimes advocated for programming at this
level. However, traditional scripts do not deal with parallelism and dataflow
aspects of workflow paradigm that makes them a limited contenders. 
%
%Add more content here to beef up the context.
%
% Goals and benefits of a workflow management system
The goals of a workflow management system for scientific computing are as
follows:

\begin{enumerate}
\item \textbf{Bulk Work.}
Workflows are often used to perform computational work in bulk. This could be
bulk data processing or running multiple distinct processing on a single set of
data. In addition to carrying out such work, workflow environments provide
oppurtunities for giving structure and order to a set of tasks and often with
sufficient expressivity, choices and selections to perform tasks selectively.

\item \textbf{User Interaction.} Adaptation to new methods of computing is not
easy for domain science users. Such users often need special assistance in
order to translate their computation requirements into concrete computing
entities.  SWEs are an ideal platform enabling users to perform many tasks
conveniently.  These tasks include composing and marshalling of execution,
browsing and selecting data, visualizing results. Furthermore,
dissemination of methods and results to communities and collaborators are
some of the activities that SWEs make convenient for end-users.

@@

% How SWEs differ from traditional programming languages and runtimes

@@

\item \textbf{Interface to Distributed Computing Infrastructures (DCIs).}
%TODO: Add content about new and emerging Distributed Computing Infrastructures: Clouds, GPUs and exascale supercomputers
Building SWE's interfaces to distributed systems is a complex task. The
distributed nature of resource utilization and asynchronous execution of
workflow requires a set of interdependent parameters to be configured
correctly. Further complexity is added because of complex dataflow
requirements, interactions with diverse tools and linking with legacy
application codes.
%
\item \textbf{Expressiveness.} Programming languages are inherently control
flow oriented where data flow is a secondary aspect. In fact, the
programmer must explicitly make sure that correct the inputs are available
to instructions and are in the right format. A program proceeds as the
execution of its statements from beginning to end and completes its
execution in a predefined sequence. However, in SWEs, the flow is often
determined by the data availability for execution. This model of
computation is known as dataflow computing. A workflow thus does not follow
a fixed order of execution and different execution sequences are possible
which are dominated by dataflow. The unit of work in SWEs is at a higher
level of granularity. A unit of execution is coarse-grained, and often
termed as ``processor'', ``task'', ``service'' or ``actor'' in the SWE
community. This notion of high-level granularity drives the depth to which
a workflow can be expressed in a given language form. In the rest of this
paper, this unit of work will be referred to as \textit{processor}. On the
other hand, a unit of execution in most programming languages is a much
simpler and finer executable statement that accomplishes a small
computational task. This presents a whole new set of requirements for
expressing the processors and how they should interact and operate upon
data.

@@

%
\begin{figure}[htb]
\begin{center}
\includegraphics[width=\linewidth]{figures/workflow_interfaces_and_language}
\caption{A generic SWE scheme showing the relationship and interaction among important components}
\label{fig:wfms}
\end{center}
\end{figure}
%
SWEs are used as tools to benefit from the underlying large-scale computing
infrastructure such as clusters, supercomputers, clouds, grids. Such
environments are in use today for production level work in different scientific
domains worldwide. They do the basic task of linking inter-related processors
of a multi-staged scientific computation. However, they differ in the features
they provide based upon the scientific domain they offer service for, and the
computational model they follow. Often, one of the main reasons for such
difference is the specialized components and requirements for a particular
domain. Owing to this fact, a given workflow system dominates a particular area
of science (eg. astronomy, biology, etc.). Nonetheless, there exist a plethora
of SWEs suitable for general purpose workflow style computations.

@@

\section{Objectives and Organization}
% What are the goals of this work?
% How is this paper supposed to help:
% -- Existing scientific users who have not adopted workflows
% -- Existing users who have adopted workflows
% -- New users
% -- Experts in the field
The present work analyzes contemporary SWEs in order to find the capabilities
and key features in terms of the characteristics described in the introduction
part of this paper. The rest of this paper is organized as follows: Section
\ref{sec:overview} presents an overview of the key SWEs studied in this work.
Section \ref{sec:scal} discusses the optimization features and resulting
scalable performance as observed in the SWEs studied. Section \ref{sec:data}
studies the data management and description aspects of the SWEs. Section
\ref{sec:inter} presents a study of SWEs approaches to interface the
large-scale computing infrastructures. Section \ref{sec:lang} details the
language features and expressiveness of the SWEs studied. Section
\ref{sec:usable} discusses the usability aspects of the SWEs.  Section
\ref{sec:sum} presents a summary and a classification of all the aspects of
SWEs studied in this work. Throughout the paper, we also present tables
summarizing the aspects of SWEs considered in respective sections. It was not
possible to obtain an exhaustive information about all the SWEs considered in
this work since we rely on the peer reviewed publications for information on a
given SWE. Some features, such as extendability of SWE strongly depend on the
extension points designed in the code; consequently, we consider it present in
only those SWEs that have a documented description or demonstration of the
given feature.

@@

Recent advances in Kepler include: MapReduce style computation, bioinformatics
computation, cloud computing~\cite{kepler2013}.

@@

%
\paragraph{Swift/Swift Script} Swift \cite{swift} is a script-based scientific
workflow environment developed at the University of Chicago. A Swift workflow
specified in the SwiftScript scripting language is translated to an
intermediate XML-based representation and enacted by the Karajan workflow
enactment engine. The SwiftScript language is proposed as a workflow language
founded on a scripting approach programmers are familiar with. To make the
language adapted to grids, the inventors of SwiftScript generalized the use of
futures \cite{Halstead85multilisp:a} for every variables defined in the script.
Futures are non-blocking assignment variables, using a proxy to ensure
immediate execution of assignment and performing lazy blocking on variable
value accesses only. The systematic use of futures for every variables in the
language makes it completely data-driven: the execution progresses
asynchronously as long as a data access is not blocking. The availability of
data enables blocked thread to restart execution. The beauty of futures is to
make this process completely hidden from the programmer who can use a scripting
language he is familiar with without ever manipulating explicit parallel
programming constructs. Swift facilitates a highly asynchronous workflow
execution. Swift supports advanced workflow constructs such as parallel loops
and conditionals. Furthermore Swift supports processing of explicitly defined
arrays of data. Swift workflows take advantage of POSIX-style shared file
system to reuse the data produced by other processors.

%Recent advances in Swift: 
1. Swift has been shown to run on multiple clouds~\cite{ccgrid2013} via its
coasters mechanism~\cite{ucc2011}.

2. Swift/T~\cite{turbine2013}, runs scientific workflows over
HPC resources. Have been shown to work on petascale supercomputers such as the
IBM BlueGene and Cray series of machines.
